---
title: "An Overview of gssr"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{An Overview of gssr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Loading the gssr package and data

```{r setup}
library(gssr)
```

As the startup message notes, the data objects are not automatically loaded. That is, we do not use R's "lazy loading" functionality. This is because the main GSS dataset is rather large. Instead we load it manually with `data()`. We'll also load the tibble that contains the codebook.

```{r load}
data(gss_all)
data(gss_doc)

gss_all

gss_doc

```

## Descriptive analysis of the data: an example

The GSS is a complex survey. When working with it, we need to take its structure into account in order to properly calculate statistics such as the population mean for a variable in some year, its standard error, and so on. For this we use the `survey` and `srvyr` packages. For details on `survey`, see Lumley (2010). We will also do some recoding, so we load several additional `tidyverse` packages to assist us.

### Packages

```{r packages}
library(dplyr)
library(ggplot2)

library(survey)
library(srvyr)
```


### Functions

Three quick-and-dirty functions, one to help clean some labels, the other to define some custom colors. 

```{r functions}
convert_agegrp <- function(x){
    x <- gsub("\\(", "", x)
    x <- gsub("\\[", "", x)
    x <- gsub("\\]", "", x)
    x <- gsub(",", "-", x)
    x <- gsub("-89", "+", x)
    regex <- "^(.*$)"
    x <- gsub(regex, "Age \\1", x)
    x
}

my_colors <- function (palette = "cb") 
{
    cb.palette <- c("#000000", "#E69F00", "#56B4E9", "#009E73", 
        "#F0E442", "#0072B2", "#D55E00", "#CC79A7")
    rcb.palette <- rev(cb.palette)
    bly.palette <- c("#E69F00", "#0072B2", "#000000", "#56B4E9", 
        "#009E73", "#F0E442", "#D55E00", "#CC79A7")
    if (palette == "cb") 
        return(cb.palette)
    else if (palette == "rcb") 
        return(rcb.palette)
    else if (palette == "bly") 
        return(bly.palette)
    else stop("Choose cb, rcb, or bly only.")
}

# from help(chartr)
capwords <- function(x, strict = FALSE) {
    cap <- function(x) paste(toupper(substring(x, 1, 1)),
                  {x <- substring(x, 2); if(strict) tolower(x) else x},
                             sep = "", collapse = " " )
    sapply(strsplit(x, split = " "), cap, USE.NAMES = !is.null(names(x)))
}

```

### Subsetting the Dataset

The GSS data retains labeling information (as it was originally imported via the `haven` package). When working with the data in an analysis, we will probably want to convert the labeled variables to data types such as factors. This should be done with care (and not on the whole dataset all at once). Typically, we will want to focus on some relatively small subset of variables and examine those. For example, let's say we want to explore the `fefam` question.

```{r setup-subset}
cont_vars <- c("year", "id", "ballot", "age")

cat_vars <- c("race", "sex", "fefam")

wt_vars <- c("vpsu",
             "vstrat",
             "oversamp",
             "formwt",              # weight to deal with experimental randomization
             "wtssall",             # weight variable
             "sampcode",            # sampling error code
             "sample")              # sampling frame and method

vars <- c(cont_vars, cat_vars, wt_vars)

gss_fam <- gss_all %>%
  select(c(cont_vars, cat_vars, wt_vars))

gss_fam

```

### Recode the Subsetted Data

Next, some recoding, along with creating some new variables.

```{r recodes}
qrts <- quantile(as.numeric(gss_fam$age), 
                 na.rm = TRUE)
qrts

quintiles <- quantile(as.numeric(gss_fam$age), 
                      probs = seq(0, 1, 0.2), na.rm = TRUE)

quintiles

## Recoding
gss_fam <- gss_fam %>%
  purrr::modify_at(vars(), zap_missing) %>%
  purrr::modify_at(wt_vars, as.numeric) %>%
  purrr::modify_at(cat_vars, as_factor) %>%
  purrr::modify_at(cat_vars, forcats::fct_relabel, capwords, strict = TRUE) %>%
  mutate(ageq = cut(x = age, breaks = unique(qrts), include.lowest = TRUE),
           ageq =  forcats::fct_relabel(ageq, convert_agegrp),
           agequint = cut(x = age, breaks = unique(quintiles), include.lowest = TRUE),
           agequint = forcats::fct_relabel(agequint, convert_agegrp),
           year_f = droplevels(factor(year)),
           young = ifelse(age < 26, "Yes", "No"),
           fefam = forcats::fct_recode(fefam, NULL = "IAP", NULL = "DK", NULL = "NA"),
           fefam_d = forcats::fct_recode(fefam,
                                Agree = "Strongly Agree",
                                Disagree = "Strongly Disagree"),
           fefam_n = recode(fefam_d, "Agree" = 0, "Disagree" = 1))

gss_fam <- gss_fam %>% 
  mutate(compwt = oversamp * formwt * wtssall, 
         samplerc = case_when(sample %in% c(3:4) ~ 3, 
                              sample %in% c(6:7) ~ 6,
                              TRUE ~ sample))

gss_fam

```

### Integrate the Survey Weights 

Now set up the survey object.

```{r weights}

options(survey.lonely.psu = "adjust")
options(na.action="na.pass")

gss_svy <- gss_fam %>%
    filter(year > 1974) %>%
    tidyr::drop_na(fefam_d, young) %>%
    mutate(stratvar = interaction(year, vstrat)) %>%
    as_survey_design(id = vpsu,
                     strata = stratvar,
                     weights = wtssall,
                     nest = TRUE)
```

## Calculate the survey-weighted means and confidence intervals for the question of interest for every year in the data

We're now in a position to calculate some properly-weighted summary statistics for the variables we're interested in. 

```{r summary}
## Get the breakdown for every year
out_ff <- gss_svy %>%
    group_by(year, sex, young, fefam_d) %>%
    summarize(prop = survey_mean(na.rm = TRUE, vartype = "ci"))

out_ff
```

### Plot the Results

We finish with a polished plot of the trends in `fefam` over time, for men and women in two (recoded) age groups over time. 

```{r fefamplot, fig.width=8, fig.height=6}
theme_set(theme_minimal())

facet_names <- c("No" = "Age Over 25 when surveyed", 
                 "Yes" = "Age 18-25 when surveyed")
fefam_txt <- "Disagreement with the statement, ‘It is much better for\neveryone involved if the man is the achiever outside the\nhome and the woman takes care of the home and family’"

out_ff %>% 
  filter(fefam_d == "Disagree") %>%
  ggplot(mapping = 
           aes(x = year, y = prop,
               ymin = prop_low, 
               ymax = prop_upp,
               color = sex, 
               group = sex, 
               fill = sex)) +
  geom_line(size = 1.2) +
  geom_ribbon(alpha = 0.3, color = NA) +
  scale_x_continuous(breaks = seq(1978, 2018, 4)) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  scale_color_manual(values = my_colors("bly")[2:1],
                     labels = c("Men", "Women"),
                     guide = guide_legend(title=NULL)) +
  scale_fill_manual(values = my_colors("bly")[2:1],
                    labels = c("Men", "Women"),
                    guide = guide_legend(title=NULL)) +
  facet_wrap(~ young, labeller = as_labeller(facet_names),
             ncol = 1) +
  coord_cartesian(xlim = c(1977, 2017)) +
  labs(x = "Year",
       y = "Percent Disagreeing",
       subtitle = fefam_txt,
       caption = "Kieran Healy http://socviz.co.\n
         Data source: General Social Survey") +
  theme(legend.position = "bottom")
```

## References

Lumley, Thomas (2010). *Complex Surveys: A Guide to Analysis Using R*. New York: Wiley.

